---
title: "TD  Week 2: Mass Shooting"
format: 
  html:
    self_contained: true
    toc: true
    embed-resources: true
execute:
  echo: true
  output: false
  warning: false
  message: false 
  error: false
editor: visual
---

# Context

The United States experiences far more mass shooting events than many other developed country in the world. While policymakers, politicians, the media, activists, and the general public recognize the widespread prevalence of these tragic events, policies intended to stop these events should be grounded in evidence and empirical data.

In July 2012, in the aftermath of a mass shooting in a movie theater in Aurora, Colorado, [Mother Jones](https://www.motherjones.com/politics/2012/07/mass-shootings-map/) published a report on mass shootings in the United States since 1982. Importantly, they provided the underlying dataset as [an open-source database](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/) for anyone interested in studying and understanding this criminal behavior.

# Instructions (read carefully!)

In this TD, you will critically analyze this dataset and identify interesting patters in the prevalence of mass shootings in the United States. Follow these guidelines:

- Submit this TD as one .qmd document and its rendered .html document on Canvas by the end of this session.

- You are **not allowed to use AI at all** throughout the duration of this TD. If you're stuck, don't hesitate to reach out to Zakia or myself

- Code must appear, but not warning messages and console outputs

- Code must be well-commented and clearly written (e.g. clear variable names, coherent spacing)

- Write properly. I don't mean grammar or orthograph, I mean: make sure that it's easy, clear and pleasant to read you. The best tip I can give is to imagine a classmate reading your paper: would they understand easily? would everything be clear?

# Exercises

## Part 1: The R Studio workflow

When you analyze data, you should make sure to centralize anything you need in one file, called the **working directory**. This is the file where you'll have your scripts, your dataset, other relevant files, and any file that is produced from your scripts (e.g. graphs, .html files rendered from Quarto...).

*Step 1: Create a working directory*

You can do this in R Studio if you click on "File", then "New Project"; then "New Directory" and "New Project". Click on "Browse" and select the file in which you want to be working. This will be your directory!

Create a working directory for this TD in your laptop file (ideally: inside a larger file dedicated to this class).

*Step 2: Set the working directory*

Using the method seen in the beginning of this class, ensure that your file is indeed defined as the current working directory. To make sure that it worked, run this code afterwards:

```{r}
getwd()
```

If you're not sure what this does, type ?getwd in the console to have more info on this function.

*Step 3: Open a .qmd file*

Go to "File", "New File", then open a Quarto document. Choose a file name but generally you can keep all default options. This is the .qmd file you will be working with during this TD.

*Step 4: Write the YAML*

The file opens with the default YAML. Replace it with the template from this .qmd file.

*Step 5: Load the dataset*

You will find a dataset called *shootings.csv* on my [Github repository for this class](https://github.com/aminelasso/SOC301B-2026), under the file called Week 1. Download it inside the working directory. Then add this chunk of code and run it:

```{r}
library(tidyverse)
data = read.csv("shootings.csv")
str(data)
```

*Step 6: Test whether "Render" works*

Click on the "Render" button above. Make sure that it works well. You should see a .html document with a title, a chunk of code, but no output.

**NB: Once you have completed this step, call Zakia or myself before moving on to the next Section.**

## Part 2: Inspect the dataset

#### Question 1:

Click on the "data" dataset in the environment on the right-hand side of R Studio. What does each row represent?

#### Question 2:

Look at the location.1 column. According to you, what does this variable tell us? Pay attention to the first rows: is there anything that we should worry about if we use this for analysis?

#### Question 3:

These are five functions (or combination of functions) that you can use to take a general look at a dataset and make crucial checks. **Memorize them**, and I recommend that you always use at least one of them before you start any analysis.

Run them all. In one sentence, explain what each one does. Which one do you think is more useful?

```{r}
str(data)
glimpse(data)
summary(data)
sum(duplicated(data)) # Looks at the number of "duplicates" in the data
colSums(is.na(data)) # Check for missing values
```

#### Question 4:

Perhaps the first question you should ask when analyzing a dataset is: how was the dataset built, and what do the variables represent? This is crucial to properly interpret the results.

First and foremost, make a quick search on the Internet: what is Mother Jones, the entity behind this dataset? Answer briefly. Why could this be important in the way we interpret the data?

#### Question 5:

Read this short piece written by the journalists who constructed this dataset: <https://www.motherjones.com/politics/2012/07/mass-shootings-map/>

*5A: How is mass shootings defined; did the definition change over time; and how does it affect data interpretation?*

*5B: What important information is lacking here, and for which you should ask for more details?*

**NB: Once you have completed this step, call Zakia or myself before moving on to the next Section.**

## Part 3: Analyzing the data

*You will find useful tutorials to answer the following questions in Chapters 2.6-2.8 of Modern Statistics with R: <https://modernstatisticswithr.com/>*

#### Question 6:

*6A:*

Generate a data frame that summarizes the number of mass shootings per year. Print the data frame as a formatted `kable()` table. *NB: To accomplish this objective, you will need to use group_by() and count().*

*6B:*

Sometimes in data analysis, we want to transform a continuous variable, like age, into a binary variable (e.g. "Child"/"Adult"). A common way to do so is to perform a *median split*.

Perform a median split on age, and print a kable function showing the mean number of victims for murderers above and below the median.

*6C:*

Sometimes in data analysis we want to *subset* the data: keep only rows from a certain category. In R, this is usually done with the *filter()* function. Write the same table as in 6B; except that it only shows the number for White people.

#### Question 7:

Generate a bar chart that identifies the number of mass shooters associated with each race category. The bars should be sorted from highest to lowest.

#### Question 8:

Generate a boxplot visualizing the number of total victims, by type of location.

#### Question 9:

Generate a plot that answers the following question: How has the total number of victims of mass shootings evolve over time ?

#### Question 10:

How does the distribution of mass shooting fatalities differ between White and Latino shooters? Add facets to your plot to show the difference in distributions.

#### Open question:

Define a research question using the criteria we learned about:

- Define a strong research question

- Define at least two credible, competing hypotheses, state which one you favor and why

- Define the prediction that each hypothesis makes on a conceptual level

- Define the same prediction on an *operational* level

- Test the prediction using the data

- Visualize the finding
